---
title: First Steps with Stan
subtitle: "Maud goes to Vegas"
author: Andrew B. Collier
location: eRum (Budapest)
date: 15 May 2018
title_logo:
- "img/logo-toptal.png"
- "img/logo-exegetic.svg"
output: 
  revealjs::revealjs_presentation:
    self_contained: false
    transition: none
    reveal_plugins: ["zoom", "notes"]
    template: datawookie-reveal-markdown-template.html
    css: datawookie-reveal-markdown.css
editor_options: 
  chunk_output_type: console
---

# {data-background="img/maud-introduction.svg" data-background-size="50%" data-background-position="left bottom"}

<!--
    Other resources:

    - http://rpubs.com/pviefers/CologneR [has interesting example of hierarchical model]
    - https://www.youtube.com/watch?v=s-9itaL1v-o "Extending the Power of R with Stan" [focus on rstanarm package]
    - https://vimeo.com/132156595
    - https://www.youtube.com/watch?v=qQFF4tPgeWI [Bob Carpenter. Background information. No demo of Stan.]
    - https://www.youtube.com/watch?v=uSjsJg8fcwY [Michael Betancourt - "Some Bayesian Modeling Techniques in Stan". Mostly linear models.]
    - https://www.youtube.com/watch?v=pHsuIaPbNbY [Michael Betancourt - "Efficient Bayesian inference with Hamiltonian Monte Carlo". Great background on MC.]

    - http://m-clark.github.io/docs/IntroBayes.html
    
    - https://matthewdharris.com/2016/10/18/estimating-a-beta-distribution-with-stan-hmc/
    - http://blackwell.math.yorku.ca/MATH6635/2016/Day17/MCMC_with_STAN_examples.html
    - https://stats.stackexchange.com/questions/165/how-would-you-explain-markov-chain-monte-carlo-mcmc-to-a-layperson
    - http://mlwhiz.com/blog/2015/08/19/MCMC_Algorithms_Beta_Distribution/
    - http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.13.7133&rep=rep1&type=pdf
    - https://jeremykun.com/2015/04/06/markov-chain-monte-carlo-without-all-the-bullshit/
    - https://link.springer.com/article/10.3758%2Fs13423-016-1015-8
-->

```{r setup, include=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.width = 7.5)

options(width = 120)

library(dplyr)
library(tidyr)
library(rstan)
library(ggplot2)
library(icon)
library(knitr)

# Plotting options for RStan.
#
rstan_gg_options(fill = "skyblue", color = "skyblue4", pt_color = "red")
```

<aside class="notes">
I'm Andrew Collier and I'm a Freelance Data Scientist at Toptal, which is a talent network for technical specialists. If anybody is interested in finding out more about Toptal, I'd be happy to chat afterwards.

Enough about me. I want you to meet somebody else.

Meet my friend Maud.

Maud is 71. She likes cats and gardening.

She also has a gambling problem.
</aside>

## {data-background="img/slot-machine.jpg"}

```{r include=FALSE}
paytable <- data.frame(
  payout = c(0, 1, 2, 5, 10, 20, 100),
  frequency = c(700, 166, 80, 33, 15, 5, 1),
  symbol = c("", rep(c("mouse", "cat"), 3)),
  repeats = c(0, 1, 1, 2, 2, 3, 3),
  stringsAsFactors = FALSE
) %>% mutate(
  symbols = purrr::map2(symbol, repeats, function(sym, n) {ifelse(sym == "", "", emo::ji(sym)) %>% rep(n) %>% paste(collapse = "")}),
  probability = frequency / sum(frequency)
) %>% select(frequency, payout, symbols)
```

<aside class="notes">
Her weakness is slot machines. She's positively obsessed by them.

To be honest, Maud is generally rather obsessive. As you'll see, this will work in our favour.
</aside>

## {data-background="img/maud-love-r.svg" data-background-size="50%" data-background-position="left bottom"}

<aside class="notes">
Somewhat surprisingly, Maud is also a big fan of R.

It's been really, um interesting, to have her as a collaborator.
</aside>

## Data (by Session)

```{r create_data, include=FALSE}
SESSIONS <- 100
AVGSPINS <- 20

# Probability of wagering 1, 2 or 3 coins per spin.
#
COINPROB = c(0.8, 0.15, 0.05)

# Seed chosen so that first session has fewer than 10 spins.
#
set.seed(538)

sessions <- tibble(
  spins = rpois(SESSIONS, AVGSPINS)
) %>% mutate(
  # Generate samples for individual spins.
  #
  details = purrr::pmap(., function(spins) {
    data.frame(
      spin = 1:spins,
      wager = sample(1:3, spins, replace = TRUE, prob = COINPROB)
    ) %>% mutate(
      payout = sample(paytable$payout, spins, TRUE, paytable$frequency) * wager
    )
  }),
  session = 1:SESSIONS
) %>% select(session, spins, details)

sessions <- inner_join(
  sessions,
  sessions %>%
    tidyr::unnest() %>%
    group_by(session) %>%
    summarise(
      hits = sum(payout > 0),
      wager = sum(wager),
      payout = sum(payout)
    )
) %>%
  mutate(
    hit_rate = hits / spins,
    rtp = payout / wager
  ) %>%
  select(session, details, everything())
```

```{r}
sessions %>% select(-details)
```

<aside class="notes">
Maud obsessively records her gambling stats. As a result we have a rather nice, rich data set to play with. At the session level the most important bits of information are the number of spins and the associated number of hits. It might occur to you that the relationship between these two quantities could be modelled as a binomial process.
</aside>

## Data (by Spin)

```{r}
sessions %>% select(session, details) %>% unnest()
```

<aside class="notes">
We can also unpack these data down to the level of individual spins, where we have something that could be modelled as a Bernoulli process.
</aside>

# Binomial Process

## {.center}

Probability distribution for binomial process:

$$
P(k | n, \theta) = \binom{n}{k} \theta^k (1 - \theta)^{n - k}
$$

where

- $k$ successes in $n$ trials where
- the probability of success on any trial is $\theta$.

<aside class="notes">
We're going to focus on the session level data for the moment. For each session we are considering the number of spins (or "trials") denoted by $n$ and the number of hits (or "successes") denoted by $k$. The probability of success on any spin is $\theta$.

The binomial distribution allows us to calculate the probability of $k$ given specific values for $n$ and $\theta$. So this assumes that you know $\theta$... But what if you don't? Certainly for most slot machines you don't know the probability of winning. If you did then you probably wouldn't be very tempted to play.

Suppose we had observations of $k$ and $n$, and wanted to estimate $\theta$.
</aside>

## {.center}

$$
\text{hit rate} = \theta^* = \frac{\text{hits}}{\text{spins}}
$$

<aside class="notes">
Well, the hit rate, $\theta$, is just the expected number of hits per spin.
</aside>

## {.center}

```{r}
with(sessions, sum(hits) / sum(spins))
```

<aside class="notes">
So we could lump all of the data together and treat it as a single massive binomial experiment, taking the ratio of the total number of hits to the total number of spins across all sessions.

This would certainly maximise the amount of data and should, in principle, give us a reasonably accurate result. But there is no way to quantify our uncertainty. Such a paucity of information leads to overconfidence in the quality of the estimate.
</aside>

## {.center}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3}
ggplot(sessions, aes(x = hit_rate)) +
  geom_histogram() +
  scale_x_continuous("Hit Rate", limits = c(0, 1), labels = scales::percent) +
  scale_y_continuous("Count") +
  theme_classic()

session_summary <- sessions %>% summarise(
  session_avg = mean(hit_rate),
  session_std = sd(hit_rate)
) %>% mutate(
  ci_lower = session_avg - 2 * session_std / sqrt(SESSIONS),
  ci_upper = session_avg + 2 * session_std / sqrt(SESSIONS)
)
```

```{r echo=FALSE}
session_summary %>% select(session_avg, session_std)
```

Confidence interval (95%) extends from `r format(session_summary$ci_lower * 100, digits = 3)`% to  `r format(session_summary$ci_upper * 100, digits = 3)`%.

<aside class="notes">
Another approach is to treat each session as an independent binomial experiment, each of which yields an estimate for the hit rate.

Now you can calculate the average and spread. However, there are two problems with this:

1. it ignores the number of spins per session (this could be incorporated by using weighted statistics) and
2. it assumes that the hit rate is Normally distributed (and it's not!).

Maud is fairly happy with this result, but it does concern her that the confidence interval on the hit rate could, in principle, extend to values less than zero or greater than one. Neither of which is physically possible. There are no constraints on $\theta$.
</aside>

---

```{r include=FALSE}
y <- c(0, 1, 0, 0, 1, 1, 0)

# Log-likelihood of a parameter value given the data.
#
bernoulli_log_likelihood <- function(theta, y) {
  sum(y * log(theta) + (1 - y) * log(1 - theta))
}

bernoulli_log_likelihood(0.3, y)
bernoulli_log_likelihood(0.6, y)
```

```{r include=FALSE}
binom_probability <- function(theta, k, n) {
  prod(choose(n, k) * theta^k * (1 - theta)^(n - k))
  #
  # Equivalent to dbinom(k, size = n, prob = theta)
}
```

For session $i$ there were $k_i$ hits from $n_i$ spins.

Assuming that sessions are independent:

$$
P(k | n, \theta) = \prod_i P(k_i | n_i, \theta) = L(\theta; n, k)
$$

<aside class="notes">
Let's consider what the probability distribution looks like for all of the sessions lumped together. If we assume that they are independent, which is a reasonable assumption, then we can simply multiply together the probability distributions for each session. This gives us, one the one hand, a distribution for the number of hits in each of the sessions. But if we think of this as a function of $\theta$ then this gives the likelihood of a particular value of $\theta$ given observed values of $k_i$ and $n_i$.

If we maximise this function then we have the most likely value of $\theta$ based on all of the data.

**Note:**

$P(k | n, \theta)$ and $L(\theta; n, k)$ are functionally equivalent but

- $P(k | n, \theta)$ is a function of $k$ (and is a probability distribution) and
- $L(\theta; n, k)$ is a function of $\theta$ (and is *not* a probability distribution).
</aside>

---

Log-likelihood for binomial process (multiple trials):

$$
LL(\theta; n, k) = \sum_i k_i \log{\theta} + (n_i - k_i) \log{(1 - \theta)}
$$

```{r echo=FALSE}
binom_likelihood <- function(theta, k, n) {
  prod(theta^k * (1 - theta)^(n - k))
}
```

```{r}
# theta - probability [single value]
# k - number of successes [vector]
# n - number of trials [vector]
#
binom_log_likelihood <- function(theta, k, n) {
  sum(k * log(theta) + (n - k) * log(1 - theta))
}
```

<aside class="notes">
Neglect the binomial coefficient since it's a constant (as a function of theta).

The log likelihood is generally easier to work with because it gets around the problem of multiplying together many small numbers.

Because the logarithm is a strictly increasing function, the log likelihood is maximised in the same place as the likelihood.
</aside>

---

```{r include=FALSE}
binom_log_likelihood(0.6, sessions$hits, sessions$spins)
binom_log_likelihood(0.6, sessions$hits, sessions$spins)

parameters <- tibble(theta = seq(0, 1, 0.02)) %>%
  mutate(
    log_likelihood = sapply(theta, binom_log_likelihood, sessions$hits, sessions$spins)
  )
```

```{r echo=FALSE, fig.height=3.5}
ggplot(parameters, aes(theta, log_likelihood)) +
  geom_point() +
  scale_y_continuous("Log likelihood")
```

```{r}
parameters %>% arrange(desc(log_likelihood)) %>% head()
```

<aside class="notes">
In the Maximum Likelihood approach we simply select the value of the parameter which gives the largest likelihood.

The assumption implicit in this approach is that in the absence of data all parameter values are equally probable.

This gives us a point estimate for the parameter. There is no indication of uncertainty. However, the likelihood is a tool which will allow us to get a much better characterisation of the hit rate.
</aside>

## Bayesian Inference

Bayes' Theorem
$$
p(\theta|y, X) = \frac{p(y|X, \theta) p(\theta)}{p(y)} \propto p(y|\theta) p(\theta)
$$
where

- $y$ are observations;
- $X$ are predictors;
- prior — $p(\theta)$ is parameter distribution <em>before</em> data;
- likelihood — $p(y|X, \theta)$ is probability of data <em>given</em> parameters; and
- posterior — $p(\theta|y, X)$ is parameter distribution <em>given</em> data.

<aside class="notes">
A Bayesian model is the combination of a prior and a likelihood.

The likelihood characterises the model and gives the probability of the data conditional on the model and choice of parameter.

The prior is specifies what we know about the model parameter before considering the data.

The denominator is sometimes called the "Evidence". It's just the marginal distribution of $y$.

It's also taken a while for these techniques to take off because they are computationally challenging. Hardware advances have made them eminently feasible.

So much for theory. Analytical expressions are rare in practice.

Confounding features:

- data are often multi-dimensional;
- models have multiple parameters.

So evaluating $p(\theta|y, X)$ becomes challenging!
</aside>

## {.center}

<blockquote>
... the theory of inverse probability is founded upon error and must be wholly rejected.
<cite>Sir Ronald Fisher (1925)</cite>
</blockquote>

<aside class="notes">
Historically the Bayesian approach has been marginalised by some eminent statisticians. For example, Ronald Fisher had this to say about inverse probability as Bayesian inference was known in his day. Evidently not everybody is a fan.

In general you can't apply Bayesian techniques analytically.
</aside>

---

```{r include=FALSE}
prior <- tibble(
  theta = seq(0, 1, 0.02),
  update = 0
)

prior_unif <- prior %>% mutate(
  p = 1
)
prior_beta <- prior %>% mutate(
  p = dbeta(theta, 8, 4)
)

binom_update <- function(prior, k, n) {
  likelihood <- sapply(prior$theta, binom_likelihood, k, n)
  #
  prior %>% mutate(
    # Update.
    p = likelihood * p,
    # Normalise.
    p = p / sum(p)
  )
}

plot(p ~ theta, prior_unif)
plot(p ~ theta, binom_update(prior_unif, 1, 3))

plot(p ~ theta, prior_beta)
plot(p ~ theta, binom_update(prior_beta, 1, 3))
```

```{r echo=FALSE}
for (n in 1:11) {
  hits = sessions[[n, "hits"]]
  spins = sessions[[n, "spins"]]
  prior_unif <- rbind(prior_unif, binom_update(prior_unif %>% filter(update == n - 1), hits, spins) %>% mutate(update = update + 1))
  prior_beta <- rbind(prior_beta, binom_update(prior_beta %>% filter(update == n - 1), hits, spins) %>% mutate(update = update + 1))
}

prior_unif$prior = "Uniform"
prior_beta$prior = "Beta"
#
priors = rbind(prior_unif, prior_beta)

ggplot(prior_unif, aes(x = theta, y = p)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~ update, scales = "free_y")
```

<aside class="notes">
One approach to applying Bayes' Theorem is to evaluate the posterior across a regular grid. Not surprisingly this is known as the "Grid Approximation".

In the grid approximation we consider a finite number of points at which we evaluate the prior and likelihood. Their product then gives us the posterior at each of those points.

There's a major problem with the grid approximation: it scales poorly with the number of parameters in the model and rapidly becomes computationally intractable.

At each iteration the model effectively learns from the new data. The distribution shifts and becomes narrower. Also note that each iteration uses the results of the previous iteration as a prior.
</aside>

---

```{r echo=FALSE}
ggplot(prior_beta, aes(x = theta, y = p)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ update, scales = "free_y")
```

---

```{r echo=FALSE}

ggplot(priors, aes(x = theta, y = p)) +
  geom_line(aes(col = prior)) +
  # geom_point() +
  facet_wrap(~ update, scales = "free_y")
```

## ... resort to Simulation!

Metropolis-Hastings algorithm.

<aside class="notes">
Draw samples from the posterior and use them to estimate quantities (like mean of parameter).

We can make this as accurate as required by generating more samples.

The only challenge is sampling from the posterior!
</aside>

## Vanilla Monte Carlo

Generate a whole bunch of independent $\theta^{(i)}$.

To do this we need to have

$$
p(\theta^{(m)} | y)
$$

<aside class="notes">
To apply Monte Carlo to a general probability distribution we'd need to apply something like importance sampling. To make this efficient we'd need to know the maximum of the distribution. This would be a significant computational burden.

Fortunately there is an alternative which is very efficient.
</aside>

## Markov Chain Monte Carlo (MCMC)

Generate $\theta^1$, $\theta^2$, $\theta^3$, ... $\theta^M$ where $\theta^m$ depends on $\theta^{m-1}$

To do this we need to have

$$
p(\theta^{m} | \theta^{m-1}, y)
$$

<aside class="notes">
We don't actually evaluate this probability explicitly. Instead we use some slight of hand.
</aside>

## Simplified Metropolis Algorithm

1. Randomly sample $\theta^{(1)}$.
2. Set $i = 2$.
3. Randomly sample proposal $\theta^{*}$ in the vicinity of $\theta^{i-1}$.
4. Sample $u$ uniformly on $[0, 1)$.
5. 

$$
\theta^{(i)} = \left\{\begin{array}{ll}
\theta^{*} & \text{if } u \cdot p(\theta^{i-1}|y, X) < p(\theta^{*}|y, X) \\
\theta^{(i-1)} & \text{otherwise.}
\end{array}\right.
$$

<aside class="notes">
So, the aim of this algorithm is to jump around in parameter space, but in a way that the probability to be at a point is proportional to the function we sample from (this is usually called the target function). In our case this is the posterior defined above.

If the probability of the new point is higher then we always accept. If it's lower then we accept with a probability that is the ratios of the densities.

The size of the "vicinity" (effectively the step size) determines the acceptance rate of new points. If it's too large then few new points will be accepted. If it's too small then the rate of diffusion will be painfully slow.

Without step 5 this would essentially give us diffusion, a random walk. But this step, the Metropolis Acceptance Procedure, ensures that sampling converges to the underlying distribution.

The principle advantage of MCMC is that we don't need to know the maximum value of the posterior.

Also it only depends on the ratio of posteriors, so these do not need to be normalised.

It might occur to you that the samples are now no longer independent. And you'd be completely correct. We'll see how this is accounted for.
</aside>

---

```{r echo=FALSE}
N = 4000

proposal <- function(theta) {
  rnorm(1, theta)
}

posterior <- function(theta) {
  dbeta(theta, 7, 4)
}

chain <- vector(mode = "numeric", length = N)
chain[1] = 0.5
#
for (n in 2:N) {
  star = proposal(chain[n-1])
  u = runif(1)
  #
  if (u < posterior(star) / posterior(chain[n-1])) {
    chain[n] = star
  } else {
    chain[n] = chain[n-1]
  }
}

chain = data.frame(
  n = 1:N,
  theta = chain
) %>% mutate(
  batch = (n - 1) %/% 1000 
)

# ggplot(chain, aes(x = theta)) +
#   geom_histogram() +
#   facet_wrap(~ batch)

ggplot(chain, aes(x = n, y = theta)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1))
```

<aside class="notes">
This is an example of a *traceplot*. You can see where the proposals are being rejected. Ideally we are aiming for good "mixing" across the domain of the distribution.

It might occur to you that the samples are now no longer independent. And you'd be completely correct. We'll see how this is accounted for.

Metropolis Monte Carlo is better than the Grid Approximation. However in an absolute sense it's not really all that good. Especially not in higher dimensions.

Metropolis sampling is rather inefficient. Stan actually uses Hamiltonian Monte Carlo. The underlying principles are essentially the same, but this technique samples the parameter space much more efficiently.

Hamiltonian Monte Carlo introduces some coherence into the random walk by using some principles from Physics. Yes, it takes a little longer to generate each sample, but those samples are guaranteed to explore parameter space much more effectively.

The NUTS sampling algorithm is *much* more efficient than vanilla MCMC.

Treat the parameter vector like the coordinates of a particle on a potential energy surface, where the potential energy is the negative log posterior.

There are more complicated schemes for proposing new locations and the rules for accepting them, but the basic idea is still: (1) pick a new "proposed" location; (2) figure out how much higher or lower that location is compared to your current location; (3) probabilistically stay put or move to that location in a way that respects the overall goal of spending time proportional to height of the location.
</aside>

## {.center}

<img src="img/logo-stan.svg" />

<aside class="notes">
Stan is a high level language for writing statistical models.
</aside>

## Stan Skeleton

<script src="https://gist.github.com/DataWookie/61d7a662ecaad87044e1c9f161a5245f.js"></script>

<aside class="notes">
The data block defines variables that have a source outside of Stan.
</aside>

## Running Stan from R

Stan workflow:

1. Data story (informs the model)
2. Write Stan program (specify likelihood and priors).
3. Stan parser converts this to C++.
4. Compile C++.
5. Execute compiled binary.
6. Evaluate results. (Possibly return to 2.)
7. Inference based on posterior sample.

Using Stan with R:

- A `.R` file. USE FONTAWESOME FILE SYMBOL!!!
- A `.stan` file. USE FONTAWESOME FILE SYMBOL!!!

<aside class="notes">
Stan is a standalone system. However it's a lot easier to run it from within R.

1. Represent a statistical model by writing its log posterior density (up to
an normalizing constant that does not depend on the unknown parameters in the
model) using the Stan modeling language. We recommend using a separate file
with a .stan extension, although it can also be done using a character string
within R.
2. Translate the Stan program to C++ code using the stanc function.
3. Compile the C++ code to create a DSO (also called a dynamic link library
(DLL)) that can be loaded by R.
4. Run the DSO to sample from the posterior distribution. This is where the magic happens.
5. Diagnose non-convergence of the MCMC chains.
6. Conduct inference based on the posterior sample (the MCMC draws from the
posterior distribution).

As mentioned earlier in the vignette, Stan programs are written in the Stan
modeling language, translated to C++ code, and then compiled to a dynamic
shared object (DSO). The DSO is then loaded by R and executed to draw the
posterior sample. The process of compiling C++ code to DSO sometimes takes a
while. When the model is the same, we can reuse the DSO from a previous run.
The stan function accepts the optional argument fit, which can be used to pass
an existing fitted model object so that the compiled model is reused. When
reusing a previous fitted model, we can still specify different values for the
other arguments to stan, including passing different data to the data
argument.

In addition, if fitted models are saved using functions like save and
save.image, RStan is able to save DSOs, so that they can be used across R
sessions. To avoid saving the DSO, specify save_dso=FALSE when calling the
stan function.
</aside>

## {.center}

```{r echo=FALSE, cache=FALSE}
read_chunk("binomial-uniform-prior.stan")
```

```{text stan_binomial_uniform_prior, message=FALSE, warning=FALSE, eval=FALSE}
```

Model with
- binomial likelihood and
- uniform prior.

<aside class="notes">
The ~ represents a stochastic relationship. An = would indicate a deterministic relationship.

If we did not specify a prior then Stan would use the uniform prior by defaulr.
</aside>

## {.center}

```{r message=FALSE, warning=FALSE, results='hide', cache=TRUE}
library(rstan)

options(mc.cores = parallel::detectCores())

trials <- list(
  N       = nrow(sessions),
  hits    = sessions$hits,
  spins   = sessions$spins
)

fit <- stan(
  file    = "binomial-no-prior.stan",
  data    = trials,
  chains  = 4,                         # Number of Markov chains
  warmup  = 1000,                      # Number of warmup iterations per chain
  iter    = 2000                       # Total number of iterations per chain
)
```

<aside class="notes">
This is the corresponding R code. Things to note:

- it loads the `rstan` library;
- sets the number of cores to be used for parallel execution (otherwise chains are run in series);
- data are transferred to Stan via a list;
- Stan is invoked via the `stan()` function; and
- parameters specify the number of chains as well as number of warmup and compute iterations per chain.

The number of Markov chains to run can be specified using the chains argument to the stan or sampling functions. By default, the chains are executed serially (i.e., one at a time) using the parent R process. There is also an optional cores argument that can be set to the number of chains (if the hardware has sufficient processors and RAM), which is appropriate on most laptops. We typically recommend first calling options(mc.cores=parallel::detectCores()) once per R session so that all available cores can be used without needing to manually specify the cores argument.
</aside>

---

```{r}
traceplot(fit, pars = c("theta"), inc_warmup = TRUE, nrow = 2)
```

---

```{r}
class(fit)

fit
```

<aside class="notes">
`n_eff` is the effective sample size. This takes into account that the samples are not independent (this is a consequence of correlation in the Markov Chain). This is a vital component of the Monte Carlo version of the Central Limit Theorem.

What's happening during the warmup phase? The parameters of the MC are getting tuned to optimally explore the parameter space (one aspect of this is choosing the step size).

Maud is immediately more comfortable with this result because, even with a uniform prior, the success rate is constrained to a reasonable domain between 0 and 1.

We don't actually get the posterior density directly. We get a bunch of samples from it instead.
</aside>

---

```{r}
list_of_draws <- extract(fit)
```

---

```{r}
df_of_draws <- as.data.frame(fit)
```

---

```{r}
fit_summary <- summary(fit)
print(names(fit_summary))
# [1] "summary"   "c_summary"
# summary - all chains merged
# c_summary - individual chains
print(fit_summary$summary)

# More specific summary.
# mu_tau_summary <- summary(fit, pars = c("mu", "tau"), probs = c(0.1, 0.9))$summary
# print(mu_tau_summary)
```

## Binomial Model: Beta Prior (Stan)

```{r echo=FALSE, cache=FALSE}
read_chunk("binomial-beta-prior.stan")
```

```{text stan_binomial_no_prior, message=FALSE, warning=FALSE, eval=FALSE}
```

<aside class="notes">
The beta distribution is the conjugate prior for the binomial likelihood (this means that the corresponding posterior will also be a beta distribution).
</aside>

---

```{r stan_binomial_beta_prior, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE, results='hide'}
library(rstan)

trials <- list(
  N       = nrow(sessions),
  hits    = sessions$hits,
  spins   = sessions$spins
)

fit <- stan(
  file    = "binomial-beta-prior.stan",
  data    = trials,
  chains  = 4,
  warmup  = 1000,
  iter    = 2000
)
```


```{r}
# Stan is generating draws from the posterior and the results (below) are based on those draws.

# print(fit, pars = c("mu", "lp__"), probs = c(0.1, 0.5, 0.9))
# print(fit)

# 80% (inner) and 95% (outer) intervals.
#
# plot(fit)
# 
# plot(fit, plotfun = "hist")
# plot(fit, plotfun = "dens")

# Autocorrelation between successive samples.
#
# plot(fit, plotfun = "ac")

# posterior_interval(fit, prob = 0.5) # Is this function from rstanarm?
#
# You can literally say that "there is a 25% chance that it's smaller than..."

pairs(fit, pars = c("theta", "lp__"), las = 1)
```


<aside class="notes">
Don't think of a confidence interval. These are quantiles of the density. Close to the median they are stable, but further into the tails they can be noisy.
</aside>

---

```{r}
plot(fit, show_density = TRUE, ci_level = 0.5, fill_color = "purple")
```
---

```{r}
plot(fit, plotfun = "hist", pars = "theta", include = FALSE)
```
---

```{r}
plot(fit, plotfun = "trace", pars = c("theta"), inc_warmup = TRUE) + ggtitle("Example of adding title to plot")
```

## Stan 3

```{r stan_poisson, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
trials <- list(
  N       = nrow(sessions),
  spins   = sessions$spins
)

fit <- stan(
  file    = "poisson.stan",
  data    = trials,
  chains  = 4,
  warmup  = 1000,
  iter    = 2000
)
```

## Stan

```{r}
extract(fit)

# hist(extract(fit)$lambda)
```

<aside class="notes">
These are samples from the posterior.
</aside>

## Stan 4

```{r cache=TRUE}
ggplot(sessions, aes(x = payout / wager)) + geom_histogram()
```

## Stan 5

```{r echo=FALSE, cache=FALSE}
read_chunk("normal.stan")
```

```{text stan_normal, message=FALSE, warning=FALSE, eval=FALSE}
```

<aside class="notes">
I know that the explicit loop is probably making you cringe. But Stan converts this into C++ code. And that code is compiled. The loop actually ends up being remarkably efficient.
</aside>

## Stan 6

```{r message=FALSE, warning=FALSE, results='hide', cache=TRUE}
trials <- list(
  N       = nrow(sessions),
  wager   = sessions$wager,
  payout  = sessions$payout
)

fit <- stan(
  file    = "normal.stan",
  data    = trials,
  chains  = 4,
  warmup  = 1000,
  iter    = 2000
)
```

## Posterior Predictive Check

## Stan 7

```{r}
pairs(fit, pars = c("mu", "sigma", "lp__"), las = 1)
```

## Regression

```{r}
ggplot(sessions, aes(spins, wager)) + geom_jitter()
```

## Regression Stan

```{r echo=FALSE, cache=FALSE}
read_chunk("regression.stan")
```

```{text stan_regression, message=FALSE, warning=FALSE, eval=FALSE}
```

<aside class="notes">
Because the number of spins has been standardised, the intercept will be the wager for the average number of spins.
</aside>

## Regression Stan R

```{r cache=TRUE, include=FALSE}
trials <- list(
  N       = nrow(sessions),
  x       = scale(sessions$spins)[,1], # Standardised covariate
  y       = sessions$wager
)

fit <- stan(
  file    = "regression.stan",
  data    = trials,
  chains  = 4,
  warmup  = 1000,
  iter    = 2000
)
```

```{r}
summary(fit)
```

<aside class="notes">
In a linear model we are saying that the response is normally distributed with a mean that depends on the covariates and a constant standard deviation. Assuming homogeneous noise.

The half-Cauchy is essentially a one-sided, heavy tailed version of the Normal distribution. This constrains the value of sigma to be positive but is not as restrictive as a Normal distribution.
</aside>

## Payline Hit Rates from RTP

```{r echo=FALSE, cache=FALSE}
read_chunk("normal-rtp.stan")
```

```{text stan_normal_rtp, message=FALSE, warning=FALSE, eval=FALSE}
```

<aside class="notes">
The priors carry a little information about the probabilities of each payline, but not a lot.
</aside>

---

```{r cache=TRUE, include=FALSE}
trials <- list(
  N       = nrow(sessions),
  rtp     = sessions$rtp
)

fit <- stan(
  file    = "normal-rtp.stan",
  data    = trials,
  chains  = 4,
  warmup  = 1000,
  iter    = 2000
)

rtp_summary = data.frame(summary(fit)$summary)
```

```{r}
summary(fit)$summary
```

The overall RTP is `r format(rtp_summary["mu", "mean"] * 100, digits = 3)`.

## Predictive Inference



## Parameter Uncertainty

Frequentist approach does not quantify uncertainty in model parameters.

## Distribution of Hit Rate

What is the distribution of the hit rate?

(Spoiler: It's not Normal.)

## Posterior Predictive Distribution

$$
p(\tilde{y}|y) = \int_\theta p(\tilde{y}|\theta) p(\theta|y) d\theta
$$

## Something

Generate some derived quantity from the posterior (because we can evaluate any function of the parameters...)

## ShinyStan

library(shinystan)
launch_shinystan(fit)
#
# Diagnose -> PPcheck "Posterior predictive check" (look at distribution of observed data versus replications - do they look similar? "If our model is a good fit then we should be able to use it to generate data that looks a lot like the data we observed.")

## What is Bayes Good For?

- Models where we care about parameter values.


<aside class="notes">
So these are not black box Machine Learning models where we simply want to make the best possible predictions. Here we actually care what's going on inside the box.
</aside>

## Resources

- http://mc-stan.org/workshops/
- <a href="http://mc-stan.org/">Stan web site</a>
- <a href="http://www.mcmchandbook.net/HandbookChapter5.pdf">MCMC Using Hamiltonian Dynamics</a> (Radford Neal)
- Blogs:
  - <http://andrewgelman.com/>
